Learning an Intervention via Causal Inference

### 2016-01-31 causal-bandit/icml-2015/general.tex
updates

We now consider single variable interventions on general causal graphs over binary variables $Y, X_1,\ldots,X_N$. 
As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of $i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we are going to assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x$. As in the previous sections we allow interventions on
a single variable by setting $X_i = j$ for any $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$. Then for $x \in \set{0,1}^N$ 
with $x_i = j$ we have
\eq{
&\P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
As before the expected reward from intervening by setting variable $X_i = j$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,,
}
where $\Omega = \set{0,1}^N$. The problem of learning the optimal intervention is now significantly complicated
by the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual model below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal, but this positive is never observed from observational samples only. This is despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\end{figure}

Since estimating $\P{Y|do(X_i=j)}$ from observational data requires conditioning on the parents of $X_i$, the variance of our estimators depends on $\P{X_i=j|\calP_i}$, rather than simply being a function of the number of times we observe $X_i=j$ as in the parallel bandit case. For example consider the graph in Figure \ref{fig:causalStructure_confounded}. In this model 
$P(Y|do(X_i = j) = \sum_z P(Y|X_i = j,Z)P(Z)$. Let $P(z=1)=.5$, $P(X_i = 1|z = 1) = 1$, and $P(X_i = 0|z = 0) = 1$ then $P(X_i = 1) = .5$ but we will be unable to estimate $P(Y|do(X_i = 1))$ as we will never get any data for the term $P(Y|X_i = 1, Z = 0)$.

For general models, it is also insufficient to only trade off learning about multiple actions by observing against learning individual actions by explicitly playing them. We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

@@

Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent,$X_k = X_{k-1}$ for $k\in {N...2}$ and $\P{X_1} = 0$. We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to explicitly play each of these arms individually to estimate their rewards.

\begin{figure}[h]
\centering
\caption{A causal chain graph}.
\label{fig:causalchain}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
\end{figure}

\todof{From here onwards needs to be replaced by truncated importance sampling version}

Let $r(x) = \P{Y = 1|X = x}$. Then

\eq {
\mu_{ij} = P(Y = 1|do(X_i = j)) = \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}r(x)
}

We wish to devise an estimator for $\mu_{ij}$ using $X$ and $Y$ without
intervening to set $X_i = j$. A natural choice is the importance weighted estimator

\eq{
Z_{ij} 
&= \frac{Y \ind{X_i = j}}{\P{X_i = j|\calP_i(X)}}\,. 
}

Checking:
\eq{
&\E{Z_{ij}} 
= \sum_x \frac{\P{X = x} r(x) \P{X = x | do(X_i = j)}}{\P{X = x}} \\
&= \P{Y = 1| do(X_i = j)} \,. 
}
The variance of $Z_{ij}$ may also be bounded
\eq{
&\Var[Z_{ij}]
\leq \E{Z_{ij}^2} \\
&\leq \sum_x \P{X = x} \left(\frac{\ind{X_i = j}}{\P{X_i = j|\calP_i(X) = \calP_i(x)}}\right)^2 \\
&= \E{\frac{1}{\P{X_i = j|\calP_i(X)}}} \\
&= V_{ij}\,.
}
In the general graph setting the dimension $m$ has a slightly different definition.
\eq{
m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}\,.
}

\todof{I think the above should be}

\eq {
m = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
}

That is, $m$ should be chosen such that the size of the set variances greater than $m$ is $m$. This ensures all the variances of the remaining arms are smaller than $m$. This definition would also seem equivalent to sorting the variances such that $V^{(0)} \geq V^{(1)} ... \geq V^{(2N)}$ and then defining $m = \min m: V^{(m)} \leq m$ which is in turn equivalent to the definition in terms of $q$ for parallel bandits.

\begin{algorithm}[H]
\caption{General Algorithm}
\begin{algorithmic}
\STATE {\bf Input:} $N$, $\mathcal P$
\STATE Compute $m$:
\eq{
m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}
}
\STATE Collect $T/2$ observational samples and estimate $\hat \mu_{ij}$:
\eq{
\hat \mu_{ij} = \sum_{t=1}^{T/2} \frac{Y_t \ind{X_{t,i} = j}}{\P{X_{t,i} = j | \mathcal P_i(X_t)}}
}
\FOR{$i \in \set{1,\ldots,m}$}
\STATE For each $i,j $ with $V_{ij} < 1/m$ take action $do(X_i = j)$ for $T/(2m)$ samples $Y^{(ij)}_1,\ldots,Y^{(ij)}_{T/2m}$
\STATE Update estimate:
\eq{
\eta &= \frac{V_{ij}}{V_{ij} + m} \\
\hat \mu_{ij} &= \eta \left(\frac{2m}{T} \sum_{t=1}^{T/(2m)} Y^{(ij)}_t\right) + (1 - \eta) \hat \mu_{ij}
}
\ENDFOR
\STATE Choose $do(X_i = j)$ for $ij = \argmax_{ij} \hat \mu_{ij}$
\end{algorithmic}
\end{algorithm}

>>>>

We now consider single variable interventions on general causal graphs over binary variables $Y, X_1,\ldots,X_N$. 
As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of $i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we are going to assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x$. As in the previous sections we allow interventions on
a single variable by setting $X_i = j$ for any $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$. Then for $x \in \set{0,1}^N$ 
with $x_i = j$ we have
\eq{
&\P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
As before the expected reward from intervening by setting variable $X_i = j$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,,
}
where $\Omega = \set{0,1}^N$. The problem of learning the optimal intervention is now significantly complicated
by the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual model below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples only, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\end{figure}

Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

@@

Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. 
If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to 
explicitly play each of these arms individually to estimate their rewards.

\begin{figure}[h]
\centering
\caption{A causal chain graph}.
\label{fig:causalchain}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
  	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
\end{figure}

Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
We would like to estimate
\eq{
\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
}
using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$.
